---
title: "Rollercoasters EDA"
output:
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: inline
---

# 1 Introduction

The objective of this report is to explore a rollercoaster dataset. We will be analyzing metrics such as rollercoaster height, speed or material of which the rollercoaster is made out of. We will try to answer some questions concerning the data, such as

1.  Who are the top 5 manufacturers by number of rollercoasters?
2.  What is the average speed and height of rollercoasters?
3.  Are newer rollercoasters faster or taller than older ones?
4.  What are the top 10 locations by number of rollercoasters?
5.  Do taller rollercoasters tend to be faster?
6.  How does the speed, height, or opening amount change over time?

The dataset used for this analysis is sourced from Kaggle (link).

# 2 Data preparation

## 2.1 Importing libraries and data

```{r}
# Importing the necessary libraries
library(tidyverse)
library(data.table)
library(dplyr)
library(GGally)
library(viridis)

# Importing the dataset
raw_data <- read.csv("coaster_db.csv")

# Setting the option for size of plots
options(repr.plot.width = 16, repr.plot.height = 9)

my_colors <- c(
    "Steel" = "#108dc7", 
    "Wood" = "#ef8e38", 
    "Other" = "#DBDBDB"
)

my_theme <- theme(
  axis.text = element_text(size = 16),  # Axis tick labels
  axis.title = element_text(size = 24),  # Axis titles
  plot.title = element_text(size = 24, face = "bold"),  # Plot title
  legend.text = element_text(size = 16),  # Legend text
  legend.title = element_text(size = 16)  # Legend title
)
```

## 2.2 Understanding the data

```{r}
# Checking the dataset dimensions
dim(raw_data)
```

```{r}
# Checking the dataset column names
colnames(raw_data)
```

```{r}
# Checking the dataset head
head(raw_data)
```

```{r}
# Checking the dataset tail
tail(raw_data)
```

## 2.3 Preparing the data

```{r}
# Selecting only the relevant columns
df <- raw_data %>%
  select(
    coaster_name, Location, Status, Type_Main, Manufacturer,
    speed_mph, height_value, height_unit, Inversions_clean, Gforce_clean,
    opening_date_clean
  )

# Checking the updated dataframe
head(df)
```

```{r}
# Creating a new unified height column
df <- df %>%
  mutate(
    Height_unified = ifelse(
      height_unit == "m", 
      height_value * 3.28084,  # Convert meters to feet
      height_value             # Keep feet as is
    )
  )
```

```{r}
# Dropping unnecessary columns
df <- df %>%
  select(-height_value, -height_unit)
```

```{r}
# Renaming columns
df <- df %>%
  rename(
    Name = coaster_name,
    Material = Type_Main,
    Speed_mph = speed_mph,
    Height_ft = Height_unified,
    Inversions = Inversions_clean,
    Gforce = Gforce_clean,
    Opening_date = opening_date_clean
  )

# Checking the updated dataframe
head(df)
```

```{r}
# Checking the structure of the dataset
str(df)
```

```{r}
# Converting `Opening_date` to Date type
df$Opening_date <- as.Date(df$Opening_date, format = "%Y-%m-%d")
```

```{r}
# Extract year from Opening_date
df <- df %>%
  mutate(Opening_year = as.numeric(format(Opening_date, "%Y")))
```

```{r}
# Summarizing basic descriptive statistics
summary(df)
```

```{r}
# Replacing empty strings with NA in all character columns
df <- df %>%
  mutate(across(where(is.character), ~ ifelse(. == "", NA, .)))

# Checking the result
head(df)
```

```{r}
# Checking for missing values in each column
colSums(is.na(df))
```

```{r}
# Replacing NA in Status and Manufacturer with "Unknown"
df <- df %>%
  mutate(
    Status = ifelse(is.na(Status), "Unknown", Status),
    Manufacturer = ifelse(is.na(Manufacturer), "Unknown", Manufacturer)
  )
```

```{r}
# Dropping the Gforce column
df <- df %>%
  select(-Gforce)
```

```{r}
# Checking for missing values
colSums(is.na(df))
```

```{r}
# Imputating missing speed_mph with the median
df <- df %>%
  mutate(Speed_mph = ifelse(is.na(Speed_mph), median(Speed_mph, na.rm = TRUE), Speed_mph))

# Imputating missing height_ft with the median
df <- df %>%
  mutate(Height_ft = ifelse(is.na(Height_ft), median(Height_ft, na.rm = TRUE), Height_ft))

# Checking for missing values
colSums(is.na(df))
```

```{r}
# Removing rows with height_ft > 500 feet
df <- df %>%
  filter(Height_ft <= 500)
```

# 3 Exploratory data analysis

## 3.1 Basic univariate analysis

```{r}
# Density plot for Speed
ggplot(df, aes(Speed_mph)) + 
    geom_density(alpha = 0.2, aes(fill = Material, colour = Material)) +
    scale_fill_manual(values = my_colors) +
    scale_color_manual(values = my_colors) +
    labs(
        title = "Density Plot of Rollercoaster Speed by Material",
        x = "Speed (mph)",
        y = "Density"
    ) +
    theme_minimal()
```

```{r}
# Histogram of speed
ggplot(df, aes(Height_ft)) + 
    geom_histogram(alpha = 0.1, binwidth = 20, aes(fill = Material, colour = Material)) +
    scale_fill_manual(values = my_colors) +
    scale_color_manual(values = my_colors) +
    labs(
        title = "Histogram of Rollercoaster Height (ft) by Material",
        x = "Height (ft)",
        y = "Density"
    ) +
    theme_minimal()
```

```{r}
# Boxplot of Height by Material
ggplot(df, aes(x = Material, y = Height_ft, fill = Material, colour = Material)) +
    geom_boxplot() +
    scale_fill_manual(values = alpha(my_colors, 0.2)) +
    scale_color_manual(values = my_colors) +
    labs(
        title = "Rollercoaster Height by Material",
        x = "Material",
        y = "Height (ft)"
    ) +
    theme_minimal()
```

```{r}
# Line Plot of Trends Over Time
df %>% count(Opening_year) %>%
    ggplot(aes(x = Opening_year, y = n)) +
    ylim(0, 40) +
    geom_line(color = "#108dc7", size = 0.5, alpha = 0.9, na.rm = TRUE) +
    geom_smooth(method = "loess", formula = "y ~ x", se = FALSE, color = "#108dc7", size = 1.5, span = 0.2, na.rm = TRUE) +
    labs(
      title = "Trends in Rollercoaster Openings Over Time", 
      x = "Year", 
      y = "Count"
    ) + 
    theme_minimal()
```

## 3.2 Basic multivariate analysis

```{r}
# Scatterplot: Speed vs. Height
ggplot(df, aes(x = Height_ft, y = Speed_mph)) +
    geom_point(alpha = 0.4, size = 3, aes(fill = Material, colour = Material)) +
    scale_fill_manual(values = my_colors) +
    scale_color_manual(values = my_colors) +
    labs(
    title = "Speed vs. Height of Rollercoasters by Material",
        x = "Height (ft)",
        y = "Speed (mph)"
    ) +
    theme_minimal()
```

```{r}
# Scatterplot: Speed vs. Inversions
ggplot(df, aes(x = Inversions, y = Speed_mph)) +
    geom_point(alpha = 0.4, size = 3, aes(fill = Material, colour = Material)) +
    scale_fill_manual(values = my_colors) +
    scale_color_manual(values = my_colors) +
    labs(
    title = "Speed vs. Inversions of Rollercoasters by Material",
        x = "Number of Inversions",
        y = "Speed (mph)"
    ) +
    theme_minimal()
```

```{r}
# Scatterplot: Height vs. Inversions
ggplot(df, aes(x = Height_ft, y = Inversions)) +
    geom_point(alpha = 0.4, size = 3, aes(fill = Material, colour = Material)) +
    scale_fill_manual(values = my_colors) +
    scale_color_manual(values = my_colors) +
    labs(
    title = "Height vs. Inversions of Rollercoasters by Material",
        x = "Height (ft)",
        y = "Number of Inversions"
    ) +
    theme_minimal()
```

# 4 Key findings

## 4.1 Who are the top 5 manufacturers by number of rollercoasters?

```{r}
# Top 5 manufacturers by number of rollercoasters
top_manufacturers <- df %>%
  group_by(Manufacturer) %>%
  summarise(Count = n()) %>%
  arrange(desc(Count)) %>%
  head(5)

# View the result
top_manufacturers
```

## 4.2 What is the average speed and height of rollercoasters?

```{r}
# Average speed and height
avg_speed <- mean(df$Speed_mph, na.rm = TRUE)
avg_height <- mean(df$Height_ft, na.rm = TRUE)

# View the results
cat("Average Speed:", round(avg_speed, 2), "mph\n")
cat("Average Height:", round(avg_height, 2), "ft\n")
```

## 4.3 Are newer rollercoasters faster or taller than older ones?

```{r}
# Trends over time
trends_over_time <- df %>%
  group_by(Opening_year) %>%
  summarise(
    Avg_Speed = mean(Speed_mph, na.rm = TRUE),
    Avg_Height = mean(Height_ft, na.rm = TRUE),
    Count = n()
  )

# Reshaping the data into long format
trends_over_time_long <- trends_over_time %>%
    pivot_longer(
        cols = c(Avg_Speed, Avg_Height),
        names_to = "Metric",
        values_to = "Value"
    )

# Plotting trends with a color legend
ggplot(trends_over_time_long, aes(x = Opening_year)) +
    geom_line(aes(y = Value, color = Metric), alpha = 0.9, size = 0.5, na.rm = TRUE) + 
    geom_smooth(aes(y = Value, color = Metric), method = "loess", formula = "y ~ x", se = FALSE, size = 1.5, span = 0.2, na.rm = TRUE) + 
    scale_color_manual(
        values = c("Avg_Speed" = "#108dc7", "Avg_Height" = "#ef8e38"), 
        labels = c("Avg_Speed" = "Speed (mph)", "Avg_Height" = "Height (ft)")
    ) +
    labs(
        title = "Trends in Rollercoaster Speed and Height Over Time",
        x = "Year",
        y = "Average Value",
        color = "Metric"
    ) +
    theme_minimal()
```

As we can clearly see from the plot, there is a positive trend in rollercoasters becoming faster and higher over time.

## 4.4 What are the top 10 locations by number of rollercoasters?

```{r}
# Top 10 locations by number of rollercoasters
top_locations <- df %>%
    filter(Location != "Other") %>%
    group_by(Location) %>%
    summarise(Count = n()) %>%
    arrange(desc(Count)) %>%
    head(10)

# View the result
top_locations
```

## 4.5 Do taller rollercoasters tend to be faster?

```{r}
# Scatterplot: Speed vs. Height
ggplot(df, aes(x = Height_ft, y = Speed_mph)) +
    geom_point(alpha = 0.4, size = 3, color = "#108dc7") +
    geom_smooth(method = "lm", formula = "y ~ x", se = TRUE, color = "#ef8e38", fill = "#ef8e38", size = 1.5, linetype = "aa") +
    labs(
    title = "Speed vs. Height of Rollercoasters",
        x = "Height (ft)",
        y = "Speed (mph)"
    ) +
    theme_minimal()
```

```{r}
# Correlation coefficient
cat("Correlation coefficient:", round(cor(df$Height_ft, df$Speed_mph, use = "complete.obs"), 2))
```

Yes, the correlation is very strong.

## 4.6 How does the amount of rollercoaster openings change over time?

```{r}
# Bar chart of Rollercoaster Openings Over Time
ggplot(df, aes(x = Opening_year)) +
    geom_bar(, color = "#108dc7", fill = "#108dc7", alpha = 0.1, na.rm = TRUE) + 
    scale_x_continuous(
        breaks = seq(1880, 2025, by = 10), 
        labels = seq(1880, 2025, by = 10)
    ) +
    labs(
        title = "Histogram of Rollercoaster Openings Over Time", 
        x = "Year", 
        y = "Count"
        ) +
    ylim(0, 40) + 
    theme_minimal() + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

As we can see from the plot, there was a peak around year 2000 and from that time there is a slight negative trend.
